% !Tex root = checkedc.tex

\parbox{5.5in}{
There are large amounts of existing critical infrastructure and
systems software written in C. This software is prone to
low-level programming errors that lead to security vulnerabilities,
correctness issues, and reliability problems.  This specification describes
Version 2 of Checked C, an extension to C designed to help software
developers improve the security, correctness, and reliability of existing
C code.  Checked C emphasizes backwards-compatibility and a
path for improving existing C code.
}

\vspace{11pt}

\parbox{5.5in}{Checked C adds 3 features: language support for
eliminating errors that cause memory corruption,
improved type-safety, and lightweight statically verifiable assertions.
These features build upon one another: avoiding memory corruption
is key to providing type-safety, and type-safety is key to being able
to statically verify C programs.  Combined, these features enable an
evidence-based approach for improving C programs.  Checked C provides a
lower-cost alternative to rewriting existing infrastructure and systems software in a type-safe
language such as Rust. Rewrites are often impractical for existing
C code: they are costly in terms of engineering effort and risk introducing
new bugs or behavior changes that cause compatibility problems.}

\vspace{11pt}

\parbox{5.5in}{
Checked C adds new pointer types and array types that are
null-checked and bounds-checked. The new pointer types
are layout-compatible with existing pointer
types, making Checked C code binary-compatible with existing C code.
Pointer bounds are described using annotations on variable
declarations that use existing program variables and data.
Bounds are checked dynamically at runtime before memory accesses.
Compiler optimizations eliminate unnecessary runtime bounds checking.
Static checking ensures the correctness of the bounds annotations.}

\vspace{11pt}

\parbox{5.5in}{Checked C also adds checked scopes. Checked scopes
provide guarantees about the bounds-safety and type-safety of code
declaration within them. In checked
scopes, only new bounds-checked and null-checked pointer types
can be used.  Type casts are also restricted to ensure type
safety.  Casts between void pointers and pointers to
 data that contains pointers are not allowed.}

\vspace{11pt}
\parbox{5.5in}{Checked C adds
generic functions and structures and existential types.
This lets most void pointer uses be replaced with type-safe code
instead.  Generic type variables are treated as incomplete types.
The restricts the use of generic type variables so that only 
single copies of functions are needed at runtime.
}
